The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-08-31 paper/paper_word_output.Rmd
Replaced all , with ; for multiple citation bibtex entries

Repeat-censused forest plots offer excellent opportunities to test neighborhood models of the effect of competition on the growth of trees [@canham_neighborhood_2004]. Neighborhood models of competition have been used to: test whether the species identity of a competitor matters [@uriarte_spatially_2004; measure species-specific competition coefficients [@das_effect_2012 @tatsumi_estimating_2016]; test competing models to see what structures competitive interactions, e.g. traits or phylogeny [@allen_permutation_2020; @uriarte_trait_2010]; and inform selective logging practices [@canham_neighborhood_2006]. Although these are well-described methods, few methods are currently available for easy application.

@@

1. Using simple feature access as the base standard for representing and encoding spatial data, rather than shapefiles [@herring_2011]. 
1. Leveraging improvements in external libraries for reading and writing spatial data (GDAL) and for geometrical operations (GEOS) [@warmerdam_geospatial_2008, @geos_2017].
1. Integrating closely with the popular `tidyverse` suite of packages for data science [@wickham_welcome_2019].

@@

We present a case-study of `forestecology`'s functionality on data from the Smithsonian Conservation Biology Institute (SCBI) large forest dynamics plot in Front Royal, VA, USA, part of the ForestGEO global network of research sites [@bourg_initial_2013, @andersonteixeira_ctfs-forestgeo_2015, @davies_forestgeo_2021]. The 25.6 ha (640 x 400 m) plot is located at the intersection of three of the major physiographic provinces of the eastern US---the Blue Ridge, Ridge and Valley, and Piedmont provinces---and is adjacent to the northern end of Shenandoah National Park.

@@

Our model of interspecific competition relies on a spatial definition of who competitor trees are: all trees within a distance `comp_dist` of a focal tree. Here we set `comp_dist` to 7.5m, a value informed by other studies [@canham_neighborhood_2004, @uriarte_spatially_2004, @canham_neighborhood_2006], but the package could also be used to compare multiple distances and see which is best supported (see Appendix \ref{mult-dist}). We use `comp_dist` and a manually constructed `sf` representation of the study region's boundary as inputs to `add_buffer_variable()` to add a `buffer` boolean variable to `growth_scbi`. All trees with `buffer` equal to `FALSE` will be our focal trees whose growth will be modeled, whereas those with `TRUE` will only act as competitor trees.

@@

The second element of spatial information we add are blocks corresponding to folds of a spatial cross-validation algorithm. Conventional cross-validation algorithms assign individual observations to folds by randomly resampling them all while assuming they are statistically independent. In the case of forest census data however, observations exhibit spatial autocorrelation. We therefore incorporate this dependence into the cross-validation algorithm by resampling spatial blocks of trees [@roberts_cross-validation_2017 @pohjankukka_estimating_2017].

@@

`run_cv()` also accepts the `run_shuffle` argument in order to permute competitor species identity as described in Section \ref{permutation-test}. Figure \ref{fig:scbi-simulation} compares model performance for 49 permutations of competitor species and RMSE calculations, both with and without cross-validation. Without cross-validation, competitor species identity does matter as the observed RMSE was significantly lower than the permutation null distribution of RMSE. However, once we incorporate spatial cross-validation, this improvement disappears. These results suggest that in this 9 ha subplot of the SCBI plot, competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other locations [@allen_permutation_2020 @uriarte_spatially_2004]. This provides a striking example of the importance of cross-validation, as without it the over-fit model gives rise to an incorrect conclusion.

@@

We also hope to extend the `forestecology` package's functionality to account for a larger variety of models for tree growth. One clear future direction would be to allow competition based on species trait values rather than species identity. There is evidence that traits predict competitive outcomes [@kunstler_competitive_2012, @lasky_trait-mediated_2014, @uriarte_trait_2010]. Thus an extension of the model would allow $\lambda$ values from Equation \ref{eq:model} to be a function of the traits of competing species.

Lastly, the `forestecology` current uses the `blockCV` package behind the scenes to create the spatial blocks acting as folds for our spatial cross-validation algorithm detailed in Sections \ref{spatial-information} and \ref{spatial-cross-validation} [@valavi_blockcv_2019]. This back-end functionality could be substituted with the `spatialsample` package for spatial resampling infrastructure; a `tidymodels` package under active development as of 2021 [@spatialsample_package, @tidymodels_package].

@@

For all the above analyses we set the cut off distance (`comp_dist`) for two stems to compete as `r comp_dist`m. This distance has been estimated in previous neighborhood competition studies in forests [@canham_neighborhood_2004, @uriarte_spatially_2004, @canham_neighborhood_2006]. We used `r comp_dist`m in @allen_permutation_2020 as an average of the values estimated in other studies. But our package can be used to find which distance is best supported by the data. Here we provide an example using another section of the SCBI plot to provide an additional example of the cross-validation block layout. To speed computation we do not consider species differences in competitive effects and treat all species as the same.

>>>>

Repeat-censused forest plots offer excellent opportunities to test neighborhood models of the effect of competition on the growth of trees [@canham_neighborhood_2004]. Neighborhood models of competition have been used to: test whether the species identity of a competitor matters [@uriarte_spatially_2004; measure species-specific competition coefficients [@das_effect_2012; @tatsumi_estimating_2016]; test competing models to see what structures competitive interactions, e.g. traits or phylogeny [@allen_permutation_2020; @uriarte_trait_2010]; and inform selective logging practices [@canham_neighborhood_2006]. Although these are well-described methods, few methods are currently available for easy application.

@@

1. Using simple feature access as the base standard for representing and encoding spatial data, rather than shapefiles [@herring_2011]. 
1. Leveraging improvements in external libraries for reading and writing spatial data (GDAL) and for geometrical operations (GEOS) [@warmerdam_geospatial_2008; @geos_2017].
1. Integrating closely with the popular `tidyverse` suite of packages for data science [@wickham_welcome_2019].

@@

We present a case-study of `forestecology`'s functionality on data from the Smithsonian Conservation Biology Institute (SCBI) large forest dynamics plot in Front Royal, VA, USA, part of the ForestGEO global network of research sites [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021]. The 25.6 ha (640 x 400 m) plot is located at the intersection of three of the major physiographic provinces of the eastern US---the Blue Ridge, Ridge and Valley, and Piedmont provinces---and is adjacent to the northern end of Shenandoah National Park.

@@

Our model of interspecific competition relies on a spatial definition of who competitor trees are: all trees within a distance `comp_dist` of a focal tree. Here we set `comp_dist` to 7.5m, a value informed by other studies [@canham_neighborhood_2004; @uriarte_spatially_2004; @canham_neighborhood_2006], but the package could also be used to compare multiple distances and see which is best supported (see Appendix \ref{mult-dist}). We use `comp_dist` and a manually constructed `sf` representation of the study region's boundary as inputs to `add_buffer_variable()` to add a `buffer` boolean variable to `growth_scbi`. All trees with `buffer` equal to `FALSE` will be our focal trees whose growth will be modeled, whereas those with `TRUE` will only act as competitor trees.

@@

The second element of spatial information we add are blocks corresponding to folds of a spatial cross-validation algorithm. Conventional cross-validation algorithms assign individual observations to folds by randomly resampling them all while assuming they are statistically independent. In the case of forest census data however, observations exhibit spatial autocorrelation. We therefore incorporate this dependence into the cross-validation algorithm by resampling spatial blocks of trees [@roberts_cross-validation_2017; @pohjankukka_estimating_2017].

@@

`run_cv()` also accepts the `run_shuffle` argument in order to permute competitor species identity as described in Section \ref{permutation-test}. Figure \ref{fig:scbi-simulation} compares model performance for 49 permutations of competitor species and RMSE calculations, both with and without cross-validation. Without cross-validation, competitor species identity does matter as the observed RMSE was significantly lower than the permutation null distribution of RMSE. However, once we incorporate spatial cross-validation, this improvement disappears. These results suggest that in this 9 ha subplot of the SCBI plot, competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other locations [@allen_permutation_2020; @uriarte_spatially_2004]. This provides a striking example of the importance of cross-validation, as without it the over-fit model gives rise to an incorrect conclusion.

@@

We also hope to extend the `forestecology` package's functionality to account for a larger variety of models for tree growth. One clear future direction would be to allow competition based on species trait values rather than species identity. There is evidence that traits predict competitive outcomes [@kunstler_competitive_2012; @lasky_trait-mediated_2014; @uriarte_trait_2010]. Thus an extension of the model would allow $\lambda$ values from Equation \ref{eq:model} to be a function of the traits of competing species.

Lastly, the `forestecology` current uses the `blockCV` package behind the scenes to create the spatial blocks acting as folds for our spatial cross-validation algorithm detailed in Sections \ref{spatial-information} and \ref{spatial-cross-validation} [@valavi_blockcv_2019]. This back-end functionality could be substituted with the `spatialsample` package for spatial resampling infrastructure; a `tidymodels` package under active development as of 2021 [@spatialsample_package; @tidymodels_package].

@@

For all the above analyses we set the cut off distance (`comp_dist`) for two stems to compete as `r comp_dist`m. This distance has been estimated in previous neighborhood competition studies in forests [@canham_neighborhood_2004; @uriarte_spatially_2004; @canham_neighborhood_2006]. We used `r comp_dist`m in @allen_permutation_2020 as an average of the values estimated in other studies. But our package can be used to find which distance is best supported by the data. Here we provide an example using another section of the SCBI plot to provide an additional example of the cross-validation block layout. To speed computation we do not consider species differences in competitive effects and treat all species as the same.
